Synthesising Recursive Functions for\\ First-Order Model Counting

### 2023-04-06 doc/talks/3_long/talk.tex
Finish the main section

\usetikzlibrary{shapes}
\usetikzlibrary{positioning}
\usetikzlibrary{overlay-beamer-styles}
\usetikzlibrary{shapes.callouts}
\usetikzlibrary{cd}
\usetikzlibrary{calc}
\usetikzlibrary{backgrounds}
\usetikzlibrary{fit}

@@

\colorlet{redish}{color2!25}
\sethlcolor{redish}
\setstcolor{black}

@@

\author{Paulius Dilkas \\ \vspace{1cm} \tiny{Joint work with Vaishak Belle (Univeristy of Edinburgh, UK)}}
\title{Synthesising Recursive Functions for\\ First-Order Model Counting}
\institute{National University of Singapore, Singapore}
\date{\today}
% TODO: update the date
% TODO: unify the use of colours
% TODO: maybe replace red!50 with one of the main colours
% TODO: \color -> \textcolor everywhere
% TODO: how do we capitalise titles of \block?
% TODO: make sure I'm using the 'forest' package whenever possible
% TODO: use Latex and not latex everywhere

@@

\begin{frame}{What Computers Can and Cannot Do}
  \begin{tikzpicture}[remember picture,overlay]
    \node[right=0.7cm] at (current page.west) (stickman) {\includegraphics[height=0.5\textheight]{stickman.png}};
    \node<-4>[left,shift=({0, -2})] at (current page.east) (computer) {\includegraphics[height=0.5\textheight]{happy_computer}};
    \node<2>[ellipse callout, draw, right=0cm, callout relative pointer={(-1, -0.5)}] at (stickman.north east) {$\int_{-\infty}^{\infty} e^{-t^{2}}\,\mathrm{d}t$};
    \node<2>[ellipse callout, draw, left=-0.5cm, below=0.5cm] at (computer.north west) {$\sqrt{\pi}$};
    \node<3>[ellipse callout, draw, right=-1cm, callout relative pointer={(-2.5, -0.5)}] at (stickman.north east) {Produce a schedule for the nurses at NUH.};
    \node<3>[ellipse callout, draw, below=0.5cm, shift=({-0.5,0}), callout relative pointer={(0.5, -0.25)}] at (computer.north west) {\includegraphics[width=0.1\textwidth]{schedule}};
    \node<4>[ellipse callout, draw, right=0cm, callout relative pointer={(-1, -0.15)}] at (stickman.north east) {Describe this picture. \includegraphics[width=0.25\textwidth]{cat.jpg}};
    \node<4>[ellipse callout, draw, below=0.5cm, shift=({-0.5,0.5}), callout relative pointer={(0.5, -0.5)}] at (computer.north west) {It's a cat with its eyes closed.};
    \node<5>[ellipse callout, draw, right=0cm, callout relative pointer={(-1, -0.15)}, text width=5cm] at (stickman.north east) {If I shuffle a deck of \alert{$n$} cards, how many possible outcomes are there?};
    \node<5>[left,shift=({-1, -1})] at (current page.east) (computer2) {\includegraphics[height=0.5\textheight]{overheating_computer}\blfootnote{Terms and conditions apply.}};
    \node[shift=({1.3, 0.1})] at (current page.south west) {\tiny \color{lightgray}{Vector art by Vecteezy.com}};
  \end{tikzpicture}
\end{frame}
% TODO: replace the cat thing with a funny DALL-E picture

@@

\begin{frame}[fragile]{(Some of the) Many Ways to Count}
  \begin{center}
    \begin{tikzcd}[remember picture]
      |[alias=WMI,visible on=<7->]|\WMI{} & |[visible on=<7->]|\WFOMI{} \\
      |[alias=WMC]|\#\SAT{}/\WMC{} \arrow[u,visible on=<7->] \ar[d,visible on=<8->] \ar[r,visible on=<5->] & |[visible on=<5->]|\textup{(W)\alert<9>{FOMC}} \arrow[u,visible on=<7->] \ar[d,visible on=<8->] \\
      |[alias=PBP,visible on=<8->]|\AMC{}, \pbp{} & |[alias=SP,visible on=<8->]|\SPPP{}
    \end{tikzcd}
  \end{center}
  \begin{tikzpicture}[overlay,remember picture]
    \begin{scope}[transform canvas={xshift=-3em}]
      \draw[->,draw=gray] ($(WMC)+(0,0.1)$)--($(WMI)+(0,0.5)$) node[midway,left,visible on=<7->] {\color{gray}{$\sum \to \int$}};
      \draw[->,draw=gray] ($(WMC)-(0,0.1)$)--($(PBP)-(0,0.5)$) node[midway,left,visible on=<8->] {\color{gray}{Generalising weights}};
    \end{scope}
    \begin{scope}[transform canvas={yshift=-2em}]
      \draw[->,draw=gray] ($(PBP)-(1,0)$)--($(SP)+(1,0)$) node[midway,below,visible on=<5->] {\color{gray}{From propositional to first-order logic}};
    \end{scope}
  \end{tikzpicture}
  \vfill
    \begin{overprint}
      \onslide<1,2>
      \begin{block}{\#SAT~\color{gray}{\parencite{DBLP:journals/tcs/Valiant79}}}
        \begin{itemize}
          \item Input formula: $\xx \lor \yy$
          \item Interpretations: $\emptyset$, $\{\, \xx \,\}$, $\{\, \yy \,\}$,
                $\{\, \xx, \yy \,\}$
          \item Models: $\{\, \xx \,\}$, $\{\, \yy \,\}$, $\{\, \xx, \yy \,\}$
          \item<2-> Answer (model count): 3
        \end{itemize}
      \end{block}
      \onslide<3,4>
      \begin{block}{Weighted Model Counting~\color{gray}{\parencite{DBLP:journals/ai/ChaviraD08}}}
        \begin{itemize}
          \item Input formula: $\xx \lor \yy$
          \item Input weights: $\begin{aligned}[t]
                                  &w(\xx) = 0.3\text{, }w(\neg \xx) = 0.7,\\
                                  &w(\yy) = 0.2\text{, }w(\neg \yy) = 0.8
                                \end{aligned}$
          \item<4-> Answer (weighted model count):
                $w(\xx)w(\yy) + w(\xx)w(\neg \yy) + w(\neg \xx)w(\yy) = 0.44$
        \end{itemize}
      \end{block}
      \onslide<5,6>
      \begin{block}{(Weighted) (Symmetric) First-Order Model Counting\\
          \color{gray}{\parencite{DBLP:conf/ijcai/BroeckTMDR11}}}
        \begin{itemize}
          \item Input formula: $\forall \xx \in \Delta\text{. }\pP(\xx)$
          \item Input weights: $w^{+}(\pP) = 0.3$, $w^{-}(\pP) = 0.7$
          \item Input domain size(s): $|\dDelta| = 2$
          \item<6-> Answer: ${(w^{+}(\pP))}^{|\dDelta|} = 0.09$
        \end{itemize}
      \end{block}
      \onslide<7>
      \begin{block}{Extensions to Continuous Domains}
        \begin{itemize}
          \item Weighted model integration
          \begin{itemize}
            \item \color{gray}{\parencite{DBLP:conf/ijcai/BellePB15}}
          \end{itemize}
          \item Weighted first-order model integration
          \begin{itemize}
            \item \color{gray}{\parencite{DBLP:conf/uai/FeldsteinB21}}
          \end{itemize}
        \end{itemize}
      \end{block}
      \onslide<8>
      \begin{block}{Generalisations of the weight function}
        \begin{itemize}
          \item Algebraic model counting
          \begin{itemize}
            \item \textcolor{gray}{\parencite{DBLP:journals/japll/KimmigBR17}}
            \item From $\mathbb{R}_{\ge 0}$ to commutative semirings
          \end{itemize}
          \item Pseudo-Boolean projection~\textcolor{gray}{\parencite{DBLP:conf/sat/DilkasB21}}
          \begin{itemize}
            \item Weights not necessarily on literals
          \end{itemize}
          \item Semiring programming~\color{gray}{\parencite{DBLP:journals/ijar/BelleR20}}
        \end{itemize}
      \end{block}
    \end{overprint}
\end{frame}
% maybe mention #CSP and other non-SAT ways of counting (SMT?, counting paths in
% automata and directed graphs)

@@

% \begin{frame}{First-Order Model Counting (in More Detail)}
%   \begin{itemize}
%     \item \alert{First-order model counting} (FOMC) is the problem of counting
%           the models of a sentence in first-order logic.
%     \item The \alert{(symmetric) weighted} variation of the problem adds weights
%           (e.g., probabilities) to predicates.
%           \begin{itemize}
%             \item It is used for efficient \alert{probabilistic inference} in
%                   relational models such as Markov logic networks.
%           \end{itemize}
%     \item None of the (implemented) (W)FOMC algorithms are able to count, e.g.,
%           \alert{injective} and \alert{bijective} functions.
%   \end{itemize}
%   \begin{block}{Claim}
%     This shortcoming can be addressed via support for \alert{recursive functions}.
%   \end{block}
% \end{frame}

@@

% TODO: it would be good to have a similar compilation walkthrough for an instance with Ref as well

\begin{frame}{Back to Our Example}
  The following function counts injections:
  \[
  f(m, n) =
  \begin{cases}
    1 & \text{if } m = 0 \text{ and } n = 0 \\
    0 & \text{if } m > 0 \text{ and } n = 0 \\
    f(m, n-1) + mf(m-1, n-1) & \text{otherwise.}
  \end{cases}
  \]
  \pause
  \begin{itemize}
    \item \alert{$f(m, n)$} can be computed in \alert{$\Theta(mn)$} time
    \begin{itemize}
      \item using dynamic programming
    \end{itemize}\pause
    \item \alert{Optimal} time complexity to compute \alert{$n^{\underline{m}}$}
          is \alert{$\Theta(m)$}
    \item But \alert{$\Theta(mn)$} is still much better than translating to
          propositional logic and running a WMC algorithm \pause
    \item The rest of this talk is about how to find such functions automatically
  \end{itemize}
\end{frame}

@@

\begin{frame}{How FCGs Are Built}
  \begin{definition}
    A \alert{(compilation) rule} is a function that takes a \alert{formula} and returns a set of \alert{$(G, L)$} pairs, where
    \begin{itemize}
    \item \alert{$G$} is an FCG,
    \item and \alert{$L$} is a list of formulas.
    \end{itemize}
  \end{definition}
  \vfill The formulas in \alert{$L$} are then \alert{compiled}, and the
  resulting FCGs are \alert{inserted} into \alert{$G$} according to a \alert{set
    order}.
\end{frame}

\begin{frame}{Example Compilation Rule: Independence}
  Input formula:
  \begin{gather}
    ({\color{color1} \forall x, y \in \Omega\text{. }x=y}) \land \label{eq:1} \tag{{\color{color1} 1}} \\
    ({\color{color2} \forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }\texttt{P}(x, y) \land \texttt{P}(x, z) \Rightarrow y=z}) \land \label{eq:2} \tag{{\color{color2} 2}} \\
    ({\color{color2} \forall w, x \in \Gamma\text{. }\forall y \in \Delta\text{. }\texttt{P}(w, y) \land \texttt{P}(x, y) \Rightarrow w=x}) \label{eq:3} \tag{{\color{color2} 3}}
  \end{gather}
  \pause
  Only one \alert{(G, L)} pair:
  \[
  G =
  \begin{tikzpicture}[edge from parent/.style={draw,-latex},baseline=(current bounding box.center)]
    \node[draw,circle] {$\land$}
    child {node {\color{color1} $\star$}}
    child {node {\color{color2} $\star$}}
    ;
  \end{tikzpicture},
  \qquad
  L = \langle {\color{color1} \eqref{eq:1}}, {\color{color2} \eqref{eq:2} \land \eqref{eq:3}} \rangle
  \]
\end{frame}
% stars mark edges to nowhere and they are associated with the formulas in L via a certain order

\begin{frame}
  \frametitle<1>{New Rule 1: Generalised Domain Recursion}
  \frametitle<2>{New Rule 2: Constraint Removal}
  \frametitle<3>{New Rule 3: Identifying Possibilities for Recursion}
  \begin{columns}
    \begin{column}{0.15\textwidth}
      \centering
      \begin{tikzpicture}[every node/.style={draw,ellipse},edge from parent/.style={draw,-latex},sibling distance=10mm,level distance=10mm]
        \node[fill=red!50, fill on=<1>] (dr) {}
        child {node {}
          child {node[fill=red!50, fill on=<2>] {}
            child {node {}
              child {node {}}
              child {node[fill=red!50, fill on=<3>] (ref) {}}
            }}};
        \draw[-latex, bend right,alt=<3>{red}{black}] (ref) to (dr);
      \end{tikzpicture}
    \end{column}
    \begin{column}{0.85\textwidth}
      \begin{overprint}
        \onslide<1>
          \begin{example}
            Input formula:
            \[
              \forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }y \ne z \Rightarrow \neg \texttt{P}(x, y) \lor \neg \texttt{P}(x, z)
            \]
            Output formula (with a new constant \alert{$c \in \Gamma$}):
            \begin{gather*}
              \forall y, z \in \Delta\text{. }y \ne z \Rightarrow \neg \texttt{P}(\alert{c}, y) \lor \neg \texttt{P}(\alert{c}, z) \\
              \begin{multlined}
                \forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }\alert{x \ne c} \land y \ne z \Rightarrow \\
                \neg \texttt{P}(x, y) \lor \neg \texttt{P}(x, z)
              \end{multlined}
            \end{gather*}
            %% Output FCG:
            %% \begin{tikzpicture}[edge from parent/.style={draw,-latex}]
            %%   \node[draw,circle] {$\textsc{DR}$}
            %%   child {node {$\star$}}
            %%   ;
            %% \end{tikzpicture}
          \end{example}
        \onslide<2>
          \begin{example}
            Input formula (with a constant \alert{$c \in \Gamma$}):
            \begin{gather*}
              \begin{multlined}
                \forall x \in \alert{\Gamma}\text{. }\forall y, z \in \Delta\text{. }\alert{x \ne c} \land y \ne z \Rightarrow \\
                \neg \texttt{P}(x, y) \lor \neg \texttt{P}(x, z)
              \end{multlined} \\
              \begin{multlined}
                \forall w,x \in \alert{\Gamma}\text{. }\forall y \in \Delta\text{. }\alert{w \ne c} \land \alert{x \ne c} \land w \ne x \Rightarrow \\
                \neg \texttt{P}(w, y) \lor \neg \texttt{P}(x, y)
              \end{multlined}
            \end{gather*}
            Output formula (with a new domain \alert{$\Gamma' \coloneqq \Gamma \setminus \{\, c \,\}$}):
            \begin{gather*}
              \forall x \in \alert{\Gamma'}\text{. }\forall y, z \in \Delta\text{. }y \ne z \Rightarrow \neg \texttt{P}(x, y) \lor \neg \texttt{P}(x, z) \\
              \forall w,x \in \alert{\Gamma'}\text{. }\forall y \in \Delta\text{. }w \ne x \Rightarrow \neg \texttt{P}(w, y) \lor \neg \texttt{P}(x, y)
            \end{gather*}
          \end{example}
        \onslide<3>
          \begin{block}{Goal}
            Check if the input formula is equivalent (up to domains) to a
            previously encountered formula.
          \end{block}
          \begin{block}{Outline}
            \begin{enumerate}
              \item Consider pairs of `similar' clauses.
              \item Consider bijections between their sets of variables.
              \item Extend each such bijection to a map between sets of domains.
              \item If the bijection makes the clauses equal, and the domain map
                    is compatible with previous domain maps, move on to another
                    pair of clauses.
            \end{enumerate}
          \end{block}
      \end{overprint}
    \end{column}
  \end{columns}
\end{frame}

>>>>

\usetikzlibrary{shapes}
\usetikzlibrary{positioning}
\usetikzlibrary{overlay-beamer-styles}
\usetikzlibrary{shapes.callouts}
\usetikzlibrary{cd}
\usetikzlibrary{calc}
\usetikzlibrary{backgrounds}
\usetikzlibrary{fit}
\usetikzlibrary{decorations.pathreplacing}

@@

\colorlet{redish}{color2!25}
\colorlet{greenish}{color1!25}
\colorlet{blueish}{color3!25}
\sethlcolor{redish}
\setstcolor{black}

@@

\author{Paulius Dilkas \\ \vspace{1cm} \tiny{Joint work with Vaishak Belle (Univeristy of Edinburgh, UK)}}
\title{Synthesising Recursive Functions for\\ First-Order Model Counting}
\institute{National University of Singapore, Singapore}
\date{18th April 2023}
% TODO: unify the use of colours, maybe replace red!50 with one of the main colours
% TODO: how do we capitalise titles of \block?
% TODO: make sure I'm using the 'forest' package whenever possible

@@

\begin{frame}{What Computers Can and Cannot Do}
  \begin{tikzpicture}[remember picture,overlay]
    \node[right=0.7cm] at (current page.west) (stickman) {\includegraphics[height=0.5\textheight]{stickman.png}};
    \node<-4>[left,shift=({0, -2})] at (current page.east) (computer) {\includegraphics[height=0.5\textheight]{happy_computer}};
    \node<2>[ellipse callout, draw, right=0cm, callout relative pointer={(-1, -0.5)}] at (stickman.north east) {$\int_{-\infty}^{\infty} e^{-t^{2}}\,\mathrm{d}t$};
    \node<2>[ellipse callout, draw, left=-0.5cm, below=0.5cm] at (computer.north west) {$\sqrt{\pi}$};
    \node<3>[ellipse callout, draw, right=0cm, callout relative pointer={(-1, -0.15)},text width=5cm] at (stickman.north east) {Produce a schedule for the nurses at the local hospital.};
    \node<3>[ellipse callout, draw, below=0.5cm, shift=({-0.5,0}), callout relative pointer={(0.5, -0.25)}] at (computer.north west) {\includegraphics[width=0.1\textwidth]{schedule}};
    \node<4>[ellipse callout, draw, right=0cm, callout relative pointer={(-1, -0.15)},text width=6cm] at (stickman.north east) {Paint a baroque oil painting of a \alert{raccoon queen wearing a crown}.};
    \node<4>[ellipse callout, draw, below=0cm, shift=({-1.5,1}), callout relative pointer={(0.4, -0.2)}] at (computer.north west) {\includegraphics[width=0.25\textwidth]{dalle.png}};
    \node<5>[ellipse callout, draw, right=0cm, callout relative pointer={(-1, -0.15)}, text width=5cm] at (stickman.north east) {If I shuffle a deck of \alert{$n$} cards, how many possible outcomes are there?};
    \node<5>[left,shift=({-1, -1})] at (current page.east) (computer2) {\includegraphics[height=0.5\textheight]{overheating_computer}\blfootnote{Terms and conditions apply.}};
    \node[shift=({1.3, 0.1})] at (current page.south west) {\tiny \textcolor{lightgray}{Vector art by Vecteezy.com}};
  \end{tikzpicture}
\end{frame}

@@

\begin{frame}[fragile]{(Some of the) Many Ways to Count}
  \begin{center}
    \begin{tikzcd}[remember picture]
      |[alias=WMI,visible on=<7->]|\WMI{} & |[visible on=<7->]|\WFOMI{} \\
      |[alias=WMC]|\#\SAT{}/\WMC{} \arrow[u,visible on=<7->] \ar[d,visible on=<8->] \ar[r,visible on=<5->] & |[visible on=<5->]|\textup{(W)\alert<9>{FOMC}} \arrow[u,visible on=<7->] \ar[d,visible on=<8->] \\
      |[alias=PBP,visible on=<8->]|\AMC{}, \pbp{} & |[alias=SP,visible on=<8->]|\SPPP{}
    \end{tikzcd}
  \end{center}
  \begin{tikzpicture}[overlay,remember picture]
    \begin{scope}[transform canvas={xshift=-3em}]
      \draw[->,draw=gray] ($(WMC)+(0,0.1)$)--($(WMI)+(0,0.5)$) node[midway,left,visible on=<7->] {\textcolor{gray}{$\sum \to \int$}};
      \draw[->,draw=gray] ($(WMC)-(0,0.1)$)--($(PBP)-(0,0.5)$) node[midway,left,visible on=<8->] {\textcolor{gray}{Generalising weights}};
    \end{scope}
    \begin{scope}[transform canvas={yshift=-2em}]
      \draw[->,draw=gray] ($(PBP)-(1,0)$)--($(SP)+(1,0)$) node[midway,below,visible on=<5->] {\textcolor{gray}{From propositional to first-order logic}};
    \end{scope}
  \vfill
    \begin{overprint}
      \onslide<1,2>
      \begin{block}{\#SAT~\textcolor{gray}{\parencite{DBLP:journals/tcs/Valiant79}}}
        \begin{itemize}
          \item Input formula: $\xx \lor \yy$
          \item Interpretations: $\emptyset$, $\{\, \xx \,\}$, $\{\, \yy \,\}$,
                $\{\, \xx, \yy \,\}$
          \item Models: $\{\, \xx \,\}$, $\{\, \yy \,\}$, $\{\, \xx, \yy \,\}$
          \item<2-> Answer (model count): 3
        \end{itemize}
      \end{block}
      \onslide<3,4>
      \begin{block}{Weighted Model Counting~\textcolor{gray}{\parencite{DBLP:journals/ai/ChaviraD08}}}
        \begin{itemize}
          \item Input formula: $\xx \lor \yy$
          \item Input weights: $\begin{aligned}[t]
                                  &w(\xx) = 0.3\text{, }w(\neg \xx) = 0.7,\\
                                  &w(\yy) = 0.2\text{, }w(\neg \yy) = 0.8
                                \end{aligned}$
          \item<4-> Answer (weighted model count):
                $w(\xx)w(\yy) + w(\xx)w(\neg \yy) + w(\neg \xx)w(\yy) = 0.44$
        \end{itemize}
      \end{block}
      \onslide<5,6>
      \begin{block}{(Weighted) (Symmetric) First-Order Model Counting\\
          \textcolor{gray}{\parencite{DBLP:conf/ijcai/BroeckTMDR11}}}
        \begin{itemize}
          \item Input formula: $\forall \xx \in \Delta\text{. }\pP(\xx)$
          \item Input weights: $w^{+}(\pP) = 0.3$, $w^{-}(\pP) = 0.7$
          \item Input domain size(s): $|\dDelta| = 2$
          \item<6-> Answer: ${(w^{+}(\pP))}^{|\dDelta|} = 0.09$
        \end{itemize}
      \end{block}
      \onslide<7>
      \begin{block}{Extensions to Continuous Domains}
        \begin{itemize}
          \item Weighted model integration
          \begin{itemize}
            \item \textcolor{gray}{\parencite{DBLP:conf/ijcai/BellePB15}}
          \end{itemize}
          \item Weighted first-order model integration
          \begin{itemize}
            \item \textcolor{gray}{\parencite{DBLP:conf/uai/FeldsteinB21}}
          \end{itemize}
        \end{itemize}
      \end{block}
      \onslide<8>
      \begin{block}{Generalisations of the weight function}
        \begin{itemize}
          \item Algebraic model counting
          \begin{itemize}
            \item \textcolor{gray}{\parencite{DBLP:journals/japll/KimmigBR17}}
            \item From $\mathbb{R}_{\ge 0}$ to commutative semirings
          \end{itemize}
          \item Pseudo-Boolean projection~\textcolor{gray}{\parencite{DBLP:conf/sat/DilkasB21}}
          \begin{itemize}
            \item Weights not necessarily on literals
          \end{itemize}
          \item Semiring programming~\textcolor{gray}{\parencite{DBLP:journals/ijar/BelleR20}}
        \end{itemize}
      \end{block}
    \end{overprint}
% maybe mention #CSP and other non-SAT ways of counting (SMT?, counting paths in
% automata and directed graphs)

@@

@@

\begin{frame}{Back to Our Example}
  The following function counts injections:
  \[
  f(m, n) =
  \begin{cases}
    1 & \text{if } m = 0 \text{ and } n = 0 \\
    0 & \text{if } m > 0 \text{ and } n = 0 \\
    f(m, n-1) + mf(m-1, n-1) & \text{otherwise.}
  \end{cases}
  \]
  \pause
  \begin{itemize}
    \item \alert{$f(m, n)$} can be computed in \alert{$\Theta(mn)$} time
    \begin{itemize}
      \item using dynamic programming
    \end{itemize}\pause
    \item \alert{Optimal} time complexity to compute \alert{$n^{\underline{m}}$}
          is \alert{$\Theta(m)$}
    \item But \alert{$\Theta(mn)$} is still much better than translating to
          propositional logic and solving a $\#\P$-complete problem \pause
    \item The rest of this talk is about how to find such functions automatically
  \end{itemize}
\end{frame}

@@

\begin{frame}{Compilation: How FCGs Are Built}
  \begin{definition}
    A \alert{(compilation) rule} is a function that takes a \alert{formula} and returns a set of \alert{$(G, L)$} pairs, where
    \begin{itemize}
    \item \alert{$G$} is an FCG,
    \item and \alert{$L$} is a list of formulas.
    \end{itemize}
  \end{definition}
  \vfill The formulas in \alert{$L$} are then \alert{compiled}, and the
  resulting FCGs are \alert{inserted} into \alert{$G$} according to a \alert{set
    order}.
\end{frame}

\begin{frame}{Example Compilation Rule: Independence}
  Input formula:
  \begin{gather}
    ({\color{color1} \forall x, y \in \Omega\text{. }x=y}) \land \label{eq:1} \tag{{\color{color1} 1}} \\
    ({\color{color2} \forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }\texttt{P}(x, y) \land \texttt{P}(x, z) \Rightarrow y=z}) \land \label{eq:2} \tag{{\color{color2} 2}} \\
    ({\color{color2} \forall w, x \in \Gamma\text{. }\forall y \in \Delta\text{. }\texttt{P}(w, y) \land \texttt{P}(x, y) \Rightarrow w=x}) \label{eq:3} \tag{{\color{color2} 3}}
  \end{gather}
  \pause
  Only one \alert{(G, L)} pair:
  \[
  G =
  \begin{tikzpicture}[edge from parent/.style={draw,-Latex},baseline=(current bounding box.center)]
    \node[draw,circle] {$\land$}
    child {node {\color{color1} $\star$}}
    child {node {\color{color2} $\star$}}
    ;
  \end{tikzpicture},
  \qquad
  L = \langle {\color{color1} \eqref{eq:1}}, {\color{color2} \eqref{eq:2} \land \eqref{eq:3}} \rangle
  \]
\end{frame}
% stars mark edges to nowhere and they are associated with the formulas in L via a certain order

\begin{frame}
  \frametitle<1>{New Rule 1/3: Generalised Domain Recursion}
  \frametitle<2>{New Rule 2/3: Constraint Removal}
  \frametitle<3>{New Rule 3/3: Identifying Possibilities for Recursion}
  \begin{columns}
    \begin{column}{0.15\textwidth}
      \centering
      \begin{tikzpicture}[every node/.style={draw,ellipse},edge from parent/.style={draw,-Latex},sibling distance=10mm,level distance=10mm]
        \node[fill=red!50, fill on=<1>] (dr) {}
        child {node {}
          child {node[fill=red!50, fill on=<2>] {}
            child {node {}
              child {node {}}
              child {node[fill=red!50, fill on=<3>] (ref) {}}
            }}};
        \draw[-Latex, bend right,alt=<3>{red}{black}] (ref) to (dr);
      \end{tikzpicture}
    \end{column}
    \begin{column}{0.85\textwidth}
      \begin{overprint}
        \onslide<1>
          \begin{example}
            Input formula:
            \[
              \forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }y \ne z \Rightarrow \neg \texttt{P}(x, y) \lor \neg \texttt{P}(x, z)
            \]
            Output formula (with a new constant \alert{$c \in \Gamma$}):
            \begin{gather*}
              \forall y, z \in \Delta\text{. }y \ne z \Rightarrow \neg \texttt{P}(\alert{c}, y) \lor \neg \texttt{P}(\alert{c}, z) \\
              \begin{multlined}
                \forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }\alert{x \ne c} \land y \ne z \Rightarrow \\
                \neg \texttt{P}(x, y) \lor \neg \texttt{P}(x, z)
              \end{multlined}
            \end{gather*}
            %% Output FCG:
            %% \begin{tikzpicture}[edge from parent/.style={draw,-latex}]
            %%   \node[draw,circle] {$\textsc{DR}$}
            %%   child {node {$\star$}}
            %%   ;
            %% \end{tikzpicture}
          \end{example}
        \onslide<2>
          \begin{example}
            Input formula (with a constant \alert{$c \in \Gamma$}):
            \begin{gather*}
              \begin{multlined}
                \forall x \in \alert{\Gamma}\text{. }\forall y, z \in \Delta\text{. }\alert{x \ne c} \land y \ne z \Rightarrow \\
                \neg \texttt{P}(x, y) \lor \neg \texttt{P}(x, z)
              \end{multlined} \\
              \begin{multlined}
                \forall w,x \in \alert{\Gamma}\text{. }\forall y \in \Delta\text{. }\alert{w \ne c} \land \alert{x \ne c} \land w \ne x \Rightarrow \\
                \neg \texttt{P}(w, y) \lor \neg \texttt{P}(x, y)
              \end{multlined}
            \end{gather*}
            Output formula (with a new domain \alert{$\Gamma' \coloneqq \Gamma \setminus \{\, c \,\}$}):
            \begin{gather*}
              \forall x \in \alert{\Gamma'}\text{. }\forall y, z \in \Delta\text{. }y \ne z \Rightarrow \neg \texttt{P}(x, y) \lor \neg \texttt{P}(x, z) \\
              \forall w,x \in \alert{\Gamma'}\text{. }\forall y \in \Delta\text{. }w \ne x \Rightarrow \neg \texttt{P}(w, y) \lor \neg \texttt{P}(x, y)
            \end{gather*}
          \end{example}
        \onslide<3>
          \begin{block}{Goal}
            Check if the input formula is equivalent (up to domains) to a
            previously encountered formula.
          \end{block}
          \begin{block}{Outline}
            \begin{enumerate}
              \item Consider pairs of `similar' clauses.
              \item Consider bijections between their sets of variables.
              \item Extend each such bijection to a map between sets of domains.
              \item If the bijection makes the clauses equal, and the domain map
                    is compatible with previous domain maps, move on to another
                    pair of clauses.
            \end{enumerate}
          \end{block}
      \end{overprint}
    \end{column}
  \end{columns}
\end{frame}

% an example on partial injection (i.e., bipartite matching) counting
\begin{frame}{How These Rules Fit Together (1/5)}
  \begin{tikzpicture}[overlay,remember picture]
    \node[draw,rounded rectangle,fill=color2!50,anchor=center,above=0.5cm] at (current page.center) (rules) {Generalised domain recursion};
    \node[above=of rules] (before) {$\begin{gathered}
        \forall x \in \alert{\Gamma}\text{. }\forall y, z \in \Delta\text{. }y \ne z \Rightarrow \neg\pP(x, y) \lor \neg\pP(x, z)\\
        \forall w,x \in \alert{\Gamma}\text{. }\forall y \in \Delta\text{. }w \ne x \Rightarrow \neg\pP(w, y) \lor \neg\pP(x, y)
      \end{gathered}$};
    \node[below=of rules] (after) {$\begin{gathered}
        \forall y, z \in \Delta\text{. }y \ne z \Rightarrow \neg\pP(\alert{c}, y) \lor \neg\pP(\alert{c}, z)\\
        \forall x \in \Gamma\text{. }\forall y,z \in \Delta\text{. }y \ne z \land \alert{x \ne c} \Rightarrow \neg\pP(x, y) \lor \neg\pP(x, z)\\
        \forall x \in \Gamma\text{. }\forall y \in \Delta\text{. }\alert{x \ne c} \Rightarrow \neg\pP(\alert{c}, y) \lor \neg\pP(x, y)\\
        \forall w \in \Gamma\text{. }\forall y \in \Delta\text{. }\alert{w \ne c} \Rightarrow \neg\pP(w, y) \lor \neg\pP(\alert{c}, y)\\
        \forall w,x \in \Gamma\text{. }\forall y \in \Delta\text{. }w \ne x \land \alert{w \ne c} \land \alert{x \ne c} \Rightarrow \neg\pP(w, y) \lor \neg\pP(x, y)
      \end{gathered}$};
    \begin{scope}[on background layer]
      \node[anchor=north,fill=greenish,minimum width=11.1cm,minimum height=0.7cm] at (before.north) {};
      \node[anchor=south,fill=blueish,minimum width=11.1cm,minimum height=0.7cm] at (before.south) {};
      \node[anchor=north,fill=greenish,minimum width=11.1cm,minimum height=1.3cm] at (after.north) {};
      \node[anchor=south,fill=blueish,minimum width=11.1cm,minimum height=2cm] at (after.south) {};
    \end{scope}
    \draw[-Latex,color=color2,ultra thick,shorten <=0.1cm,shorten >=0.1cm] (before) -- (rules);
    \draw[-Latex,color=color2,ultra thick,shorten <=0.1cm,shorten >=0.1cm] (rules) -- (after);
  \end{tikzpicture}
\end{frame}

\begin{frame}{How These Rules Fit Together (2/5)}
  \begin{tikzpicture}[overlay,remember picture]
    \node[draw,rounded rectangle,fill=color2!50,anchor=center,below=1cm] at (current page.center) (rules) {Atom counting and unit propagation};
    \node[above=of rules] (before) {$\begin{gathered}
        \forall y, z \in \Delta\text{. }y \ne z \Rightarrow \neg\pP(\alert{c}, y) \lor \neg\pP(\alert{c}, z)\\
        \forall x \in \Gamma\text{. }\forall y,z \in \Delta\text{. }y \ne z \land \alert{x \ne c} \Rightarrow \neg\pP(x, y) \lor \neg\pP(x, z)\\
        \forall x \in \Gamma\text{. }\forall y \in \Delta\text{. }\alert{x \ne c} \Rightarrow \neg\pP(\alert{c}, y) \lor \neg\pP(x, y)\\
        \forall w \in \Gamma\text{. }\forall y \in \Delta\text{. }\alert{w \ne c} \Rightarrow \neg\pP(w, y) \lor \neg\pP(\alert{c}, y)\\
        \forall w,x \in \Gamma\text{. }\forall y \in \Delta\text{. }w \ne x \land \alert{w \ne c} \land \alert{x \ne c} \Rightarrow \neg\pP(w, y) \lor \neg\pP(x, y)
      \end{gathered}$};
    \node[below=of rules] (after) {$\begin{gathered}
        \forall y, z \in \alert{\Delta^{\top}}\text{. }y \ne z \Rightarrow \bot\\
        \forall x \in \Gamma\text{. }\forall y, z \in \alert{\Delta^{\bot}}\text{. }y \ne z \land x \ne c \Rightarrow \neg\pP(x, y) \lor \neg\pP(x, z) \\
        \forall w, x \in \Gamma\text{. }\forall y \in \alert{\Delta^{\bot}}\text{. }w \ne x \land w \ne c \land x \ne c \Rightarrow \neg\pP(w, y) \lor \neg\pP(x, y)
      \end{gathered}$};
    \draw[-Latex,color=color2,ultra thick,shorten <=0.1cm,shorten >=0.1cm] (before) -- (rules);
    \draw[-Latex,color=color2,ultra thick,shorten <=0.1cm,shorten >=0.1cm] (rules) -- (after);
    \draw[-Latex,color=color2,ultra thick,shorten <=0.1cm,shorten >=0.1cm] (rules) -- (after);
  \end{tikzpicture}
\end{frame}
